---
title: "COVID-19 Mortality and Testing Rates by Country"
author: "Karen Nershi"
date: "3/19/2020"
output: html_document
---

```{r, include=FALSE, warning=FALSE, error=FALSE, message=FALSE}

```



Recently, I read [Andreas Backhaus's great post](https://medium.com/@andreasbackhausab/coronavirus-why-its-so-deadly-in-italy-c4200a15a7bf) about the age distribution of COVID-19 cases across South Korea and Italy. In short, South Korea shows a higher proportion of young people with the virus, with 30% of the nation's cases between the ages of 20 and 29. Italy, meanwhile, has a distribution of cases much more heavily skewed toward the elderly, which does not mirror the broader population. This difference can help explain why South Korea has a vastly different mortality rate than Italy -- just under 1% compared to Italy's 8.3% based on the [latest data](https://coronavirus.jhu.edu/map.html). But why is there a vastly diffferent age distribution across COVID-19 cases between the two countries?

One clear reason is the difference in testing between the two countries. South Korea has tested more capita -- 5,567 tests per million -- compared to Italy's 2,514 tests per million. This shows that South Korea has carried out 2.1 times as many tests relative to its population as Italy, which affects both countries' mortality rates. Since South Korea has tested more widely and shows a broader distribution of cases across age groups, this suggests that Italy's outbreak actually includes many more young people who are not currently represented in the official data. This could explain why Italy's mortality rate appears high compared to other countries', as the denominator of official cases likely does not include a large number of young people who show mild symptoms.

### Global Testing

Although not all countries publish data on the number of COVID-19 cases they test, [Our World in Data](https://ourworldindata.org/coronavirus) provides data for the ones that do.The data shows a great deal of variation across countries, ranging from 2.55 tests per million in Pakistan to 27,888 tests per million in Kuwait. The United States has tested relatively few cases so far, with 64 tests for every million citizens. The map below shows the number of COVID-19 tests per million for available countries.
```{r, include=FALSE}
load("df.RData")

library(rworldmap)

mapped_data <- joinCountryData2Map(df, joinCode = "ISO3", nameJoinColumn = "ccode")
```


```{r cars, echo=FALSE, message=F, warning=F, error=F}

par(mai=c(.5,0,1,0),xaxs="i",yaxs="i")
mapParams <- mapCountryData(mapped_data, 
                            nameColumnToPlot = "Total.COVID.19.tests.per.million.people",
                            mapTitle = "COVID-19 Tests per Million",
                            addLegend = F,
                            borderCol = "black") 

do.call(addMapLegend, c(mapParams, legendLabels="all",
                        legendWidth=0.5,
                        legendIntervals="data",
                        legendShrink=0.6,
                        legendMar=3))
```

In Europe, which currently has the highest number of COVID-19 cases globally, Iceland, Norway, and Sweden have conducted the most per capita tests at over 3,000 tests per million, while Ukraine, Finland, Hungary, and France have all conducted fewer than 170 tests per million.

```{r, include=FALSE}
load("small.RData")

library(rworldmap)

mapped_data <- joinCountryData2Map(small, joinCode = "ISO3", nameJoinColumn = "ccode")
```


```{r, echo=F, output=F}
mapParams <- mapCountryData(mapped_data, 
                            nameColumnToPlot = "Total.COVID.19.tests.per.million.people",
                            mapTitle = "COVID-19 Tests per Million in Europe",
                            addLegend = F,
                            borderCol = "black",
                            mapRegion = 'europe') 

do.call(addMapLegend, c(mapParams, legendLabels="all",
                        legendWidth=0.5,
                        legendIntervals="data",
                        legendShrink=0.6,
                        legendMar=2))
```


### Confirmed Cases 

Confirmed cases, meanwhile, are highly concentrated in Europe. Iceland and Italy top the chart with 6,637 and 2,514 cases per million respectively. What does the relationship between tested aand confirmed cases across countries tell us?

```{r, include=FALSE}
mapped_data <- joinCountryData2Map(df, joinCode = "ISO3", nameJoinColumn = "ccode")
```


```{r echo=F, message=FALSE, warning=FALSE}
par(mai=c(.5,0,1,0),xaxs="i",yaxs="i")

mapParams <- mapCountryData(mapped_data, 
                            nameColumnToPlot = "Total.confirmed.cases.of.COVID.19.per.million.people",
                            mapTitle = "COVID-19 Confirmed Cases per Million",
                            addLegend = F,
                            borderCol = "black") 

do.call(addMapLegend, c(mapParams, legendLabels="all",
                        legendWidth=0.5,
                        legendIntervals="data",
                        legendShrink=0.6,
                        legendMar=3))
```

### Rate of Confirmed Cases
One way to consider this relationship is look at what percentage of tested cases are confirmed as positive cases. Italy had the highest percentage of confirmed cases from those tested (18.8%), followed by the Philippines (18.1%), France (12.7%), Denmark (12.6%), and Austria (11.0%). The United States ranked 20 out of the 48 countries, with 4.0% of all tested cases confirmed COVID-19 cases. South Korea, which has been widely praised for its response to the crisis, had 2.9% of all tested cases confirmed as active cases. This rate is displayed in the map below.

```{r, include=FALSE}
load('ratio.RData')

library(rworldmap)

mapped_data <- joinCountryData2Map(rat, joinCode = "ISO3", nameJoinColumn = "ccode")
```


```{r, echo=FALSE, warning=FALSE}
par(mai=c(.5,0,1,0),xaxs="i",yaxs="i", cex=1)

suppressWarnings(mapParams <- mapCountryData(mapped_data, 
                            nameColumnToPlot = "rate_confirmed",
                            mapTitle = "Rate of Confirmed Cases for Tested COVID-19 Cases",
                            addLegend = F,
                            borderCol = "black") )

suppressWarnings(do.call(addMapLegend, c(mapParams, legendLabels="all",
                        legendWidth=0.5,
                        legendIntervals="data",
                        legendShrink=0.6,
                        legendMar=3)))
```

Interpreting this rate, however, is tricky. A country could have a very high rate of confirmed cases if it is not testing widely, which would indicate many more undiagnosed cases for each confirmed case. On the other hand, countries may test all individuals in contact with sick people, in which case a high rate would signal a high level of effectiveness (i.e., countries find and test sick people). Without knowing each country's testing strategy, it is difficult to understand what this rate means for how countries are testing for COVID-19.

### What does this mean?

Since many COVID-19 patients show mild or moderate symptoms, the more widely a country tests, the lower the mortality rate will appear. It is also important to note that it takes a while for COVID-19 to kill patients, so countries that are early in the pandemic will have an artificially low mortality rate since they will have many diagnosed cases but few deaths... *yet.* Accordingly, this data does not tell the whole story as it compares countries at different stages, and a truly fair comparison would compare mortality and testing rates after the pandemic has run its course.

```{r, echo=F, warning=F, error=F, message=F}
load('full.RData')

library(ggplot2)
library(ggrepel)
ggplot(full, aes(full$mortality_rate, full$Total.COVID.19.tests.per.million.people,
                        label=Entity.x)) + 
  geom_smooth() +
  geom_point() +
  geom_label_repel(aes(label = Entity.x),
                   box.padding   = 0.35, 
                   point.padding = 0.5,
                   segment.color = 'grey50') +
  theme_classic() + 
  xlab("Mortality Rate") + 
  ylab("COVID-19 Tests per Million") +
  ggtitle('Tests per Million vs. Mortality Rate') + 
  theme(plot.title = element_text(hjust = 0.5))

```

That being said, we discover the expected relationship -- countries that test more appear to have lower mortality rates, like South Korea. Again, this is likely a statistical issue more than anything. Although some individuals have suffered adverse health outcomes because they were not diagnosed early in the illness, at the national level, it is more likely that countries with lower levels of testing appear to have higher mortality rates because people with milder symptoms are not included in the official statistics.

Finally, the quality of a nation's healthcare system also plays an important role in determining both mortality rates and testing. However, for the moment, I am overlooking thiss as many of the countries with high numbers of COVID-19 cases today are Western countries with developed healthcare systems. Countries also differ in median age, which affects mortality rates, and may differ in other important ways not captured here. However, this graph gives an idea of the relationship between testing and mortality rates.

This data suggests that countries testing at a lower rate may show higher mortality rates. The wide variation in testing also suggests that there can be significant difference in the types of COVID-19 cases a country has despite similar reported numbers. This all suggests that there are significant differences in both the amount and type of testing countries perform, which helps explain diverging mortality rates across countries.